---
title: "Analysis2.6_Figure_plotting"
author: "AndrewHart"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Description: This Script is both exploratory and functions to produce many figures. First, MAST based DEG analyss was performed and the results analyzed through a variety of mehcanisms. Next, UCell enrichment analysis of gene sets was performed and analyzed 
# Figures produced in this script include Figure S5A, Figure 4A, Figure 4B, Figure 4C, Figure 3C , Figure 3D, Figure 3F, Figure 3H, Figure S4E, Figure 3E, Figure 3G , Figure S4A, Figure S4B, Figure S4C, Figure S4D
#Load Libraries 
```{r}
library(Seurat)
library(Signac)
library(dplyr)
library(ggplot2)
library(reshape2)
library(data.table)
library(cowplot)
library(scCustomize)
library(forcats)
library(ggbreak) 
library(presto) 
library(ggforce)
library(SingleR)
library(celldex)
library(harmony)
#library(Azimuth)
library(glmGamPoi) 
library(future)
library(dsb) 
library(SeuratDisk) 
library(RColorBrewer)
library(monocle3)
library(gt)
library(SeuratWrappers)
library(pheatmap)
library(grid)
library(gridExtra)
library(qs2)
library(tidyr)
library(tibble)
library(ggfortify)
library(ggrepel)
library(purrr)
options(future.globals.maxSize = 5000 * 1024^2)
```

#Establish Directories
```{r} 
getwd() #/home/hartandrew
setwd("/path/to/analysis/directory")
out <- "/path/to/analysis/directory/Output"
version <- "/path/to/analysis/directory/Version"
seurat <- "/path/to/analysis/directory/Seurat_Files"
images <- "/path/to/analysis/directory/Images"
CSV <- "/path/to/analysis/directory/CSV"

```
#Set colors
```{r}
colors7 <- c("#980339", "#379FFB", "#D6A206", "#027102", "#6313B7", "#AFAFAF","#542600" )
colors6 <- c("#980339", "#379FFB", "#D6A206", "#027102", "#6313B7", "#AFAFAF")
colors5 <- c("#980339", "#379FFB", "#D6A206", "#027102", "#6313B7")
colors8 <- c("#332288", "#88CCEE", "#117733", "#A27DF1", "#DDCC77",  "#AA4499", "#CC6677", "#882255")
colors9 <- c("#332288", "#88CCEE", "#117733","#44AA99", "#A27DF1", "#DDCC77",  "#AA4499", "#CC6677", "#882255")
cbf_18_colors <- c(
  "#1170AA",  
  "#FCB13F",  
  "#60BFC1",  
  "#EF6F6A",  
    "#937860", 
  "#D17A00",  
  "#B78CBA",  
  "#B3B3B3",  
    "#64B200", 
  "#D45E00",  
  "#7E8CCF",  
  "#E6A0C4",  
  "#568B3F",  
  "#C44E52",
   "#5FA2A3", 
  "#CCB974", 
  "#D0A6BE", 
  "#4E84C4"  
)
```

#Load Objects
```{r}
#MLN_filt <- qs_read("/path/to/analysis/directory/Seurat_Files/Analysis2_MLN_clustered_annotated.qs2")
Ileum_filt <- qs_read("/path/to/analysis/directory/Seurat_Files/Analysis2.5_Ileum.qs2")
```

# Remove the Duodenum from the data  we focus on the ileum to contrast between infections
```{r}
unique(Ileum_filt$SampleType)
Idents(Ileum_filt) <- "SampleType"
Ileum_filt_i <- subset(Ileum_filt, idents = c("Ileum Lamina Propria" , "Ileum Intestinal Epithelial Cells" ))
cell_counts <- Ileum_filt_i@meta.data %>%
  group_by(InfectionStatus, Mouse, secondlevel) %>%
  summarise(n_cells = n(), .groups = "drop")

# For each InfectionStatus and secondlevel, count how many mice have >=10 cells
qualified_celltypes <- cell_counts %>%
  group_by(InfectionStatus, secondlevel) %>%
  summarise(mice_with_10plus = sum(n_cells >= 10), .groups = "drop") %>%
  filter(mice_with_10plus < 2)

# Get list of problematic cell types
bad_celltypes <- qualified_celltypes %>%
  distinct(secondlevel) %>%
  pull(secondlevel)

# View result
print("Cell types with <10 cells in at least 2 mice for any condition:")
print(bad_celltypes)
length(unique(Ileum_filt_i$secondlevel))

```

# Compare Each Cell Type for Each condition - A separate script was run to perform the analysis in base R - DEG_MAST.r
# The above script produces the DEG output read in below

```{r}
library(tidyr)
# MAST method for DEG analysis takes a very long time
all_conditions <- setdiff(unique(Ileum_filt$InfectionStatus), "Naive")
all_celltypes <- unique(Ileum_filt$secondlevel)

all_degs_df <- data.table::fread("/path/to/analysis/directory/CSV/all_degs_df_20250918.csv")

results_list <- list()
sig_degs <- all_degs_df %>% filter(global_padj < 0.05)

# Summarize DEGs per condition × cell type
summary_df <- sig_degs %>%
  group_by(condition, celltype) %>%
  summarise(
    total_deg = n(),
    up_deg = sum(avg_log2FC > 0),
    down_deg = sum(avg_log2FC < 0),
    avg_abs_logFC = mean(abs(avg_log2FC), na.rm = TRUE),
    .groups = "drop"
  )
# Fill in missing entries
for (cond in all_conditions) {
  for (ct in all_celltypes) {
    key <- paste(cond, ct, sep = "_")
    if (!key %in% names(results_list)) {
      # Add from summary_df if present
      matched <- summary_df %>% filter(condition == cond & celltype == ct)
      if (nrow(matched) > 0) {
        results_list[[key]] <- matched
      } else {
        results_list[[key]] <- tibble(
          condition = cond,
          celltype = ct,
          total_deg = 0,
          up_deg = 0,
          down_deg = 0
        )
      }
    }
  }
}

results_df <- bind_rows(results_list)
# Remove the cell types which did not have at least 10 cells in Naive for comparison 
results_df_filt <- results_df[!results_df$celltype %in% c("cDC1s", "Cycling ILCs", "Mast Cells", "Mesothelial Cells", "Neutrophils"),]
```



```{r}
#This chunk makes a bubble plot where each bubble is a pie chart showing the number of upregulated and downregulated genes for ecah cell types within each infection from DEG analysis 
cellorder <- c("Enteric Glia",  "Monocytes" , "Macrophages" , "Plasmacytoid DCs" ,   "cDC2s" , 
                                    
               "ILC1s" , "ILC2s" ,  "ILC3s"  ,   "NK Cells"  , "NKT Cells" ,
               "Cycling T Cells",  "CD4 ab T Cells" , "Effector CD4 ab T Cells" , "CD8 ab T Cells" , "Effector CD8 ab T Cells" ,    "gd T Cells",
                "GC B Cells" ,  "B Cells", "Plasma Cells", 
                  "Stem Cells" , "Transit Amplifying Cells" ,  "Enterocytes" ,  "Goblet Cells"  ,  "Tuft Cells"   ,  "Paneth Cells"   ,  "Enteroendocrine Cells" ,
                "Lymphatic Endothelial Cells" , "Blood Endothelial Cells",   "Fibroblasts",         "Smooth Muscle Cells"   , "Stromal Cells"     )

infections <- c("Yersinia", "SFB_YA", "Nippostrongylus", "MNV", "Cryptosporidium", "Candida" )

all_combos <- expand.grid(
  condition = infections,
  celltype = cellorder,
  stringsAsFactors = FALSE
)

x_spacing <- 4   # Spread bubbles apart horizontally
y_spacing <- 4 

results_df_filt$celltype <- factor(results_df_filt$celltype, levels = cellorder)
results_df_filt$condition <- factor(results_df_filt$condition, levels = infections)
plot_df <- full_join(all_combos, results_df_filt, by = c("condition", "celltype")) %>%
  replace_na(list(total_deg = 0, up_deg = 0, down_deg = 0)) %>%
  mutate(
    total = up_deg + down_deg,
    up_frac = ifelse(total > 0, up_deg / total, NA),
    down_frac = ifelse(total > 0, down_deg / total, NA),
    start_up = 0,
    end_up = 2 * pi * up_frac,
    start_down = end_up,
    end_down = 2 * pi,
    # Axis positions
    x = as.numeric(factor(condition, levels = infections)) * x_spacing,
    y = as.numeric(factor(celltype, levels = cellorder)) * y_spacing,
    # Scaled bubble radius using log10
    bubble_radius = log10(total + 1) / 2 
  )


p <- ggplot() +
  #light gray background circle for tested + 0 DEGs
  geom_point(
    data = plot_df %>% filter(total == 0),
    aes(x = y, y = x),
    shape = 21, fill = "lightgray", color = "black", size = 3
  ) +
  
  # Upregulated pie slice
  geom_arc_bar(
    data = plot_df %>% filter(total > 0),
    aes(
      x0 = y, y0 = x,
      r0 = 0, r = bubble_radius,
      start = start_up, end = end_up,
      fill = "Upregulated"
    ),
    color = "black"
  ) +
  
  # Downregulated pie slice
  geom_arc_bar(
    data = plot_df %>% filter(total > 0),
    aes(
      x0 = y, y0 = x,
      r0 = 0, r = bubble_radius,
      start = start_down, end = end_down,
      fill = "Downregulated"
    ),
    color = "black"
  ) +
  scale_x_continuous(
    breaks = sort(unique(plot_df$y)),
    labels = cellorder) +
  scale_y_continuous(
    breaks = sort(unique(plot_df$x)),
    labels = infections) +
  scale_fill_manual(values = c("Upregulated" = "#99000D", "Downregulated" = "#08306B")) +
  coord_fixed() + 
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(), axis.title = element_blank(),
    legend.position = "right"
  ) +
  labs( x = "Condition", y = "Cell Type", fill = "Direction of DEGs")


print(p)
ggsave( "CEllTypexCondition_DEG_BubblePie_Nolegend_sep2025.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 13,  height =17,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)
```



```{r}
#Here I will perform two analyses. I will look only at DEGs regardless of cell type and create PCA and similarity indices based on this information across conditions


deg_table_all <- all_degs_df %>%
  filter(global_padj < 0.05)  # Or your preferred DEG threshold

# Create binary gene x condition matrix (1 = DE in any cell type)
binary_mat <- deg_table_all %>%
  distinct(condition, gene) %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = condition, values_from = value, values_fill = 0) %>%
  column_to_rownames("gene")

# Jaccard distance matrix - calculates overlap in DEGs between any two conditions, regardless of cell type
library(vegan)
jaccard_dist <- vegdist(t(binary_mat), method = "jaccard")

# Hierarchical clustering
hc <- hclust(jaccard_dist, method = "average")
plot(hc, main = "Clustering of Conditions by DEG Gene Overlap (Jaccard)", xlab = "")

binary_mat_filtered <- binary_mat[rowSums(binary_mat) > 0, ]  # remove all-0s
binary_mat_filtered <- binary_mat_filtered[apply(binary_mat_filtered, 1, function(x) var(x) > 0), ]  # remove constant rows

#Remove conditions with no variation
binary_mat_filtered <- binary_mat_filtered[, apply(binary_mat_filtered, 2, function(x) var(x) > 0) ]

pca_res <- prcomp(t(binary_mat_filtered), scale. = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_df$condition <- rownames(pca_df)
ggplot(pca_df, aes(x = PC1, y = PC2, label = condition)) +
  geom_point(size = 4, color = "#2c7fb8") +
  geom_text_repel(size = 5, max.overlaps = 100) +
  theme_minimal(base_size = 14) +
  labs(title = "PCA of DEG Gene Overlap Between Conditions",
       x = paste0("PC1 (", round(summary(pca_res)$importance[2, 1] * 100, 1), "% variance)"),
       y = paste0("PC2 (", round(summary(pca_res)$importance[2, 2] * 100, 1), "% variance)"))


#This is the second half of the analysis and it includes information about the cell types - so in order for a gene to provide meaningful interaction or links between conditions, it must be DE in the same cell type. 

deg_filtered <- all_degs_df %>%
  filter(global_padj < 0.05 & avg_log2FC >1)
deg_filtered <- deg_filtered %>%
  mutate(gene_celltype = paste(gene, celltype, sep = "__"))
binary_mat_gc <- deg_filtered %>%
  distinct(gene_celltype, condition) %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = condition, values_from = value, values_fill = 0) %>%
  column_to_rownames("gene_celltype")

binary_mat_gc <- binary_mat_gc[apply(binary_mat_gc, 1, function(x) var(x) > 0), ]
binary_mat_gc <- binary_mat_gc[, apply(binary_mat_gc, 2, function(x) var(x) > 0) ]
# Transpose rows = conditions, columns = gene__celltype features
pca_gc <- prcomp(t(binary_mat_gc), scale. = TRUE)

# Get PCA coordinates 
pca_gc_df <- as.data.frame(pca_gc$x)
pca_gc_df$condition <- rownames(pca_gc_df)

ggplot(pca_gc_df, aes(x = PC1, y = PC2, label = condition)) +
  geom_point(size = 4, color = "#2c7fb8") +
  geom_text_repel(size = 5) +
  theme_minimal(base_size = 14) +
  labs(title = "PCA of DEG Overlap (Gene + Cell Type Specific)",
       x = paste0("PC1 (", round(summary(pca_gc)$importance[2, 1] * 100, 1), "% variance)"),
       y = paste0("PC2 (", round(summary(pca_gc)$importance[2, 2] * 100, 1), "% variance)"))

ggplot(pca_gc_df, aes(x = PC3, y = PC2, label = condition)) +
  geom_point(size = 4, color = "#2c7fb8") +
  geom_text_repel(size = 5) +
  theme_minimal(base_size = 14) +
  labs(title = "PCA of DEG Overlap (Gene + Cell Type Specific)",
       x = paste0("PC3 (", round(summary(pca_gc)$importance[2, 3] * 100, 1), "% variance)"),
       y = paste0("PC2 (", round(summary(pca_gc)$importance[2, 2] * 100, 1), "% variance)"))



# Make Small Multiples to visualize the differences in gene expression  
scores_df <- as.data.frame(pca_gc$x[,1:3])
scores_df$condition <- rownames(scores_df)

melted <- reshape2::melt(scores_df,
                         id.vars = "condition",
                         variable.name = "PC",
                         value.name = "score")
ggplot(melted, aes(x = condition, y = score, fill = condition)) +
  geom_col() +
  facet_wrap(~ PC, scales = "free_y") + coord_flip() +
  theme_minimal()



library(vegan)
jaccard_gc <- vegdist(t(binary_mat_gc), method = "jaccard")
hc_gc <- hclust(jaccard_gc, method = "average")
plot(hc_gc, main = "Hierarchical Clustering: Gene x Cell Type DEG Overlap", xlab = "")

library(ggdendro)
dend_data <- ggdendro::dendro_data(hc_gc)
ggplot() +
  geom_segment(data = segment(dend_data),
               aes(x = x, y = y, xend = xend, yend = yend),
               linewidth = 0.6) +
  geom_text(data = label(dend_data),
            aes(x = x, y = y, label = label),
            hjust = 0, 
            size = 4,
            fontface = "bold") +
  coord_flip() +  
  theme_minimal(base_size = 14) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.margin = margin(10, 50, 10, 10) 
  ) +
  labs(title = "Hierarchical Clustering: DEG Overlap by Condition", x = NULL, y = NULL)

ggsave(paste0(images, "/Dendogram_DEG_similarity_byCondition.svg"), width = 4, height = 8, plot = last_plot())

```


```{r}
# Create a bar plot which shows the total number of upregulated and downregulated genes across all cell types in and across infections

all_degs_df <- all_degs_df %>%
  filter(global_padj < 0.05)

# Separate upregulated and downregulated genes
up_genes <- all_degs_df %>%
  filter(avg_log2FC > 0) %>%
  group_by(condition) %>%
  summarise(
    unique_up_genes = n_distinct(gene),
    .groups = "drop"
  )

down_genes <- all_degs_df %>%
  filter(avg_log2FC < 0) %>%
  group_by(condition) %>%
  summarise(
    unique_down_genes = n_distinct(gene),
    .groups = "drop"
  )

# Combine into one summary table
deg_summary_by_condition <- full_join(up_genes, down_genes, by = "condition") %>%
  arrange(condition)


deg_summary_long <- deg_summary_by_condition %>%
  pivot_longer(cols = c(unique_up_genes, unique_down_genes),
               names_to = "Direction", values_to = "Count") %>%
  mutate(
    Direction = case_when(
      Direction == "unique_up_genes" ~ "Upregulated",
      Direction == "unique_down_genes" ~ "Downregulated"
    ),
    Count = ifelse(Direction == "Downregulated", -Count, Count)
  )

deg_summary_long$condition <- factor(deg_summary_long$condition, levels = infections)

ggplot(deg_summary_long, aes(x = Count, y = condition, fill = Direction)) +
  geom_bar(stat = "identity", position = "identity") +
  scale_fill_manual(values = c("Upregulated" = "#99000D", "Downregulated" = "#08306B")) +
  labs(
    x = "Number of DEGs"
  ) +
  theme_minimal() +
  geom_vline(xintercept = 0, color = "black") +
  theme(
    text = element_text(size = 14),
    axis.title.y = element_blank(),
    legend.position = "none"
  )
ggsave( "Combined_DEGS-across cellTypes_Barplot_sep2025.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 6,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)
```


```{r}
# Examine both the number and trends in overall strength (logFC) of DEGs across the different cell types. Violin plot

sig_degs <- all_degs_df %>%
  filter(global_padj < 0.05)

# Summarize average log2FC per gene per condition
avg_lfc_by_condition <- sig_degs %>%
  group_by(condition) %>%
  summarise(
    mean_log2FC = mean(avg_log2FC, na.rm = TRUE),
    .groups = "drop"
  )

# Each condition will have a "line" of gene points sorted by logFC

# Line Chart
sig_degs <- sig_degs %>%
  mutate(
    signed_logfc = sign(avg_log2FC) * log10(abs(avg_log2FC) + 1)
  )

ggplot(sig_degs, aes(x = signed_logfc, y = ..density.., color = condition)) +
  geom_density(size = 1) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Signed log10-transformed avg_log2FC",
    y = "Density",
    color = "Condition"
  )


# Plot frequency polygon
ggplot(sig_degs, aes(x = signed_logfc, color = condition)) +
  geom_freqpoly(binwidth = 0.05, size = 1.2) +  
  theme_minimal(base_size = 14) +
  scale_color_manual(values = cbf_18_colors) +
  labs(
    x = "Log10 transformed Avg LogFC",
    y = "Number of genes",
    color = "Infection"
  ) +
  theme(
    panel.grid.major = element_line(color = "grey90"),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_text(size = 14), 
    panel.grid.minor = element_blank(), 
    axis.text = element_text(color = "black")
  )
ggsave( "LineChart_of_DEG_by_Infection_Fig4_oct2025.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 6,  height = 4,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)


# Make sure infection (condition) is ordered consistently
sig_degs$condition <- factor(
  sig_degs$condition,
  levels = c("Naive", "Candida", "Cryptosporidium", "MNV", 
             "Nippostrongylus", "SFB_YA", "Yersinia")
)

sig_degs <- sig_degs %>%
  mutate(
    signed_logfc = sign(avg_log2FC) * log10(abs(avg_log2FC) + 1)
  )
ggplot(sig_degs, aes(x = condition, y = signed_logfc, fill = condition)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.8, color = "black", linewidth = 0.2) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5, color = "black") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = cbf_18_colors) +
  labs(
    x = NULL,
    y = "Log10 LogFC",
    fill = "Condition",
    title = "Distribution of DE Gene Expression Changes per Infection"
  ) +
  theme(
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    legend.position = "none"
  )

#Figure S5A produced here
# Save
ggsave(
  "Violin_DEG_logFC_by_Infection_Fig4_oct2025_transformed.svg",
  plot = last_plot(),
  device = NULL,
  path = images,
  width = 6,
  height = 9,
  units = "in",
  dpi = 600,
  limitsize = TRUE,
  bg = NULL
)
```

```{r}
# create a Barplot for Enterocytes/StemCells/TA showing the total number of DEGs in these cell types across the infections
all_degs_df_entero <- all_degs_df %>%
  filter(celltype %in% c("Enterocytes", "Stem Cells", "Transit Amplifying Cells")) %>%
  filter(global_padj < 0.05 )


# Separate upregulated and downregulated genes
up_genes_entero <- all_degs_df_entero %>%
  filter(avg_log2FC > 0) %>%
  group_by(condition) %>%
  summarise(
    unique_up_genes = n_distinct(gene),
    .groups = "drop"
  )

down_genes_entero <- all_degs_df_entero %>%
  filter(avg_log2FC < 0) %>%
  group_by(condition) %>%
  summarise(
    unique_down_genes = n_distinct(gene),
    .groups = "drop"
  )

# summary table
deg_summary_by_condition_e <- full_join(up_genes_entero, down_genes_entero, by = "condition") %>%
  arrange(condition)

deg_summary_long_entero <- deg_summary_by_condition_e %>%
  pivot_longer(cols = c(unique_up_genes, unique_down_genes),
               names_to = "Direction", values_to = "Count") %>%
  mutate(
    Direction = case_when(
      Direction == "unique_up_genes" ~ "Upregulated",
      Direction == "unique_down_genes" ~ "Downregulated"
    ),
    Count = ifelse(Direction == "Downregulated", -Count, Count)
  )

deg_summary_long_entero$condition <- factor(deg_summary_long_entero$condition, levels = infections)

ggplot(deg_summary_long_entero, aes(x = Count, y = condition, fill = Direction)) +
  geom_bar(stat = "identity", position = "identity") +
  scale_fill_manual(values = c("Upregulated" = "#99000D", "Downregulated" = "#08306B")) +
  labs(
    x = "Number of DEGs"
  ) +
  theme_minimal() +
  geom_vline(xintercept = 0, color = "black") +
  theme(
    text = element_text(size = 14),
    axis.title.y = element_blank(),
    legend.position = "none"
  )

ggsave( "Combined_DEGS_Enterocytes_Stem_Transit_Barplot_sep2025.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 6,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)

```

```{r}
#Create a stacked bar chart of all the DEGs and the cell types they occur in  - this shows us numerically which cell types have the most DEGs
# First looking only at upregulated DEGs
#Figure 4A and Figure 4B produced here 
Upregulated <- all_degs_df %>%
  filter(  global_padj < 0.05 & abs(avg_log2FC) > 0)

deg_counts <- Upregulated %>%
  group_by(celltype, condition) %>%
  summarise(n_genes = n_distinct(gene), .groups = "drop")

# Reorder cell types by total n_genes across conditions
deg_counts <- deg_counts %>%
  group_by(celltype) %>%
  mutate(total_genes = sum(n_genes)) %>%
  ungroup() %>%
  mutate(celltype = reorder(celltype, -total_genes))  

ggplot(deg_counts, aes(x = celltype, y = n_genes, fill = condition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "DEGs per Cell Type by Condition",
    x = "Cell Type",
    y = "Number of DEGs"
  ) +
  scale_fill_manual(values = cbf_18_colors) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.x = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(), 
    text = element_text(fill = "black")
  )
ggsave( "Stacked_BAr_DEGCounts_per_CellType_and_Infection_total.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 8,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)


Upregulated <- all_degs_df %>%
  filter( avg_log2FC > 0, global_padj < 0.05)

deg_counts <- Upregulated %>%
  group_by(celltype, condition) %>%
  summarise(n_genes = n_distinct(gene), .groups = "drop")

# Reorder cell types by total n_genes across conditions
deg_counts <- deg_counts %>%
  group_by(celltype) %>%
  mutate(total_genes = sum(n_genes)) %>%
  ungroup() %>%
  mutate(celltype = reorder(celltype, -total_genes))  # negative for descending order

ggplot(deg_counts, aes(x = celltype, y = n_genes, fill = condition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Upregulated DEGs per Cell Type by Condition",
    x = "Cell Type",
    y = "Number of DEGs"
  ) +
  scale_fill_manual(values = cbf_18_colors) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.x = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(), 
    text = element_text(fill = "black")
  )
  #Figure 4A
ggsave( "Stacked_BAr_DEGCounts_per_CellType_and_Infection_upregulated.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 8,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)


Upregulated <- all_degs_df %>%
  filter( avg_log2FC > 0.583, global_padj < 0.05)

deg_counts <- Upregulated %>%
  group_by(celltype, condition) %>%
  summarise(n_genes = n_distinct(gene), .groups = "drop")

# Reorder cell types by total n_genes across conditions
deg_counts <- deg_counts %>%
  group_by(celltype) %>%
  mutate(total_genes = sum(n_genes)) %>%
  ungroup() %>%
  mutate(celltype = reorder(celltype, -total_genes))  # negative for descending order

ggplot(deg_counts, aes(x = celltype, y = n_genes, fill = condition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Upregulated (logfc) DEGs per Cell Type by Condition",
    x = "Cell Type",
    y = "Number of DEGs"
  ) +
  scale_fill_manual(values = cbf_18_colors) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.x = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(), 
    text = element_text(fill = "black")
  )
  #Figure 4B
ggsave( "Stacked_BAr_DEGCounts_per_CellType_and_Infection_upregulated_LogFC.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 8,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)

Downregulated <- all_degs_df %>%
  filter( avg_log2FC < 0, global_padj < 0.05)

deg_counts <- Downregulated %>%
  group_by(celltype, condition) %>%
  summarise(n_genes = n_distinct(gene), .groups = "drop")

deg_counts <- deg_counts %>%
  group_by(celltype) %>%
  mutate(total_genes = sum(n_genes)) %>%
  ungroup() %>%
  mutate(celltype = reorder(celltype, -total_genes)) 

ggplot(deg_counts, aes(x = celltype, y = n_genes, fill = condition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Downregulated DEGs per Cell Type by Condition",
    x = "Cell Type",
    y = "Number of DEGs"
  ) +
  scale_fill_manual(values = cbf_18_colors) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.x = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(), 
    text = element_text(fill = "black")
  )
ggsave( "Stacked_BAr_DEGCounts_per_CellType_and_Infection_downregulated.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 8,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)
```



```{r}
# A Bar chart of total DEGs in ecah infection - not split by up/down regulated
total <- all_degs_df %>%
  filter(global_padj <0.05) %>%
  group_by(condition) %>%
  summarise(
    unique_genes = n_distinct(gene),
    .groups = "drop"
  )

ggplot(total, aes(x = unique_genes, y = condition), fill  = "black") +
  geom_bar( stat = "identity", position = "identity") +
  labs(
    x = "Number of DEGs"
  ) +
  theme_minimal() +
  geom_vline(xintercept = 0, color = "black") +
  theme(
    text = element_text(size = 14),
    axis.title.y = element_blank(),
    legend.position = "none"
  )

ggsave( "Total_DEG_across_Infection_BArchart_sep2025.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 6,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)
```




#Upset plot and ontology----
```{r}
#Upset plot showing shared genes among those upregulated during infections across all cell types
library(ComplexUpset)

unique_deg <- all_degs_df %>%
  filter(global_padj < 0.05 ) %>%
  dplyr::select(gene, condition) %>%
  distinct()
nrow(unique_deg[unique_deg$condition == "Cryptosporidium"])
nrow(unique_deg[unique_deg$condition == "Yersinia"])
nrow(unique_deg[unique_deg$condition == "MNV"])
nrow(unique_deg[unique_deg$condition == "Nippostrongylus"])
nrow(unique_deg[unique_deg$condition == "SFB_YA"])
nrow(unique_deg[unique_deg$condition == "Candida"])


up_degs_df <- all_degs_df %>%
  filter(global_padj < 0.05 & avg_log2FC > 0) %>%
  dplyr::select(gene, condition) %>%
  distinct()

up_matrix <- up_degs_df %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = condition, values_from = present, values_fill = FALSE)

up_mat <- as.data.frame(up_matrix)
rownames(up_mat) <- up_mat$gene
up_mat$gene <- NULL


upset(
  up_mat,
  intersect = colnames(up_mat),
  name = "Upregulated Genes",
  width_ratio = 0.2,
  base_annotations = list(
    'Intersection size' = intersection_size(
      text = list(
        angle = 90,
        hjust = -0.5,
        vjust = 0.5,
        size = 3.5)) + theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )))


ggsave( "Upset_upregulated_DEGS_celltype_comparisons_sep2025.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 12,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)
```

```{r}
#Save all the overlap combinations  - the up_mat has rownames as genes and colnames as conditions and is a logical matrix for whether each gene (row) is found within each cross comparison

condition_names <- colnames(up_mat)
#Create a list of all combinations 
combinations <- unlist(lapply(1:length(condition_names), function(x) {
  combn(condition_names, x, simplify = FALSE)
}), recursive = FALSE)

# empty excel workbook
wb <- openxlsx::createWorkbook()

for (i in seq_along(combinations)) {
  combo <- combinations[[i]]
  combo_name <- paste(combo, collapse = "_AND_") # By providing the index name at the beginning I can make sure all tab names are unique
  
  sheet_name <- substr(paste0(i, "_", combo_name), 1, 31)
  
   genes_in_combo <- rownames(up_mat)[apply(up_mat[, combo, drop = FALSE], 1, all) &
    rowSums(up_mat[, setdiff(condition_names, combo), drop = FALSE]) == 0]
  
  if (length(genes_in_combo) > 0) {
    addWorksheet(wb, sheetName = sheet_name)
    
    df_out <- data.frame(
      Gene = genes_in_combo,
      Comparison = paste(combo, collapse = ", "),
      stringsAsFactors = FALSE
    )
    
    writeData(wb, sheet = sheet_name, x = df_out)
  }
}

saveWorkbook(wb, paste0(CSV, "/Upregulated_Overlap_Genes.xlsx"), overwrite = TRUE)
```


```{r}
#The enterocyte development line make up the biggest chunks of DEGS in the different infections and these are interesting cells - we will focus specifically on these populations and examine the different up set plots showing how genes are arranged

up_degs_df_e <- all_degs_df_entero %>%
  filter(global_padj < 0.05, avg_log2FC >  0) %>%
  select(gene, condition) %>%
  distinct()

up_matrix_e <- up_degs_df_e %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = condition, values_from = present, values_fill = FALSE)

up_mat_e <- as.data.frame(up_matrix_e)
rownames(up_mat_e) <- up_mat_e$gene
up_mat_e$gene <- NULL


upset(
  up_mat_e,
  intersect = colnames(up_mat_e),
  name = "Upregulated Genes Entero/Stem/Transit",
  width_ratio = 0.2,
  base_annotations = list(
    'Intersection size' = intersection_size(
      text = list(
        angle = 90,
        hjust =-0.5,
        vjust = 0.5,
        size = 3.5)) + theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )))
#Figure 4C
ggsave( "Upset_upregulated_DEGS_Entero_Stem_Transit_comparisons.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 12,  height = 5,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)


```

# Enrichment analysis was performed utilizing a separate script and results were loaded as below

#Enrichment Results 
```{r}
# Load in the results from the Enrichment scoring and add as metadata to the object
hallmark <- data.table::fread("/path/to/analysis/directory/CSV/Hallmark_ssGSEA_UCell.csv")
rownames(hallmark) <- hallmark$cell_id
all(rownames(hallmark) %in% colnames(Ileum_filt))
hallmark <- hallmark[,-1] # remove cell id column 
genesets_hallmark <- colnames(hallmark)
Ileum_filt@meta.data <- Ileum_filt@meta.data %>%
  dplyr::select(-starts_with("HALLMARK_"))
Ileum_filt@assays[genesets_hallmark] <- NULL
Ileum_filt <- AddMetaData(Ileum_filt, metadata = hallmark)

gs_IL27 <- data.table::fread("/path/to/analysis/directory/CSV/Il27_Ucell.csv")
gs_IL27b <- gs_IL27
rownames(gs_IL27) <- gs_IL27$cell_id
all(rownames(gs_IL27) %in% colnames(Ileum_filt))
gs_IL27 <- gs_IL27[,-1]
genesets_gs_IL27 <- colnames(gs_IL27)
Ileum_filt@meta.data <- Ileum_filt@meta.data %>%
  dplyr::select(-starts_with("gs_IL27_"))
Ileum_filt@assays[genesets_gs_IL27] <- NULL
rownames(gs_IL27) <- gs_IL27b$cell_id
Ileum_filt <- AddMetaData(Ileum_filt, metadata = gs_IL27)

Th2 <- data.table::fread("/path/to/analysis/directory/CSV/Th2_ssGSEA_Ucell.csv")
Th17 <- data.table::fread("/path/to/analysis/directory/CSV/Th17_ssGSEA_Ucell.csv")
rownames(Th2) <- Th2$cell_id
rownames(Th17) <- Th2$cell_id
all(rownames(Th2) %in% colnames(Ileum_filt))
Th2 <- Th2[,-1] 
genesets_Th2 <- colnames(Th2)

Ileum_filt <- AddMetaData(Ileum_filt, metadata = Th2)

all(rownames(Th17) %in% colnames(Ileum_filt))
Th17 <- Th17[, -1] 
Th2 <- data.table::fread("/path/to/analysis/directory/CSV/Th2_ssGSEA_Ucell.csv")
rownames(Th17) <- Th2$cell_id
genesets_Th17 <- colnames(Th17)

Ileum_filt <- AddMetaData(Ileum_filt, metadata = Th17)

```


```{r}
#Create a data frame of the score of selected Pathways 
pathways_interest  <- c("HALLMARK_INTERFERON_ALPHA_RESPONSE","HALLMARK_BILE_ACID_METABOLISM" , 
         "HALLMARK_GLYCOLYSIS" , 
         "HALLMARK_INFLAMMATORY_RESPONSE", 
         "HALLMARK_WNT_BETA_CATENIN_SIGNALING", 
         "HALLMARK_FATTY_ACID_METABOLISM" ,
         "HALLMARK_IL2_STAT5_SIGNALING" , 
         "HALLMARK_TGF_BETA_SIGNALING" ,
         "HALLMARK_IL6_JAK_STAT3_SIGNALING" , 
         "HALLMARK_OXIDATIVE_PHOSPHORYLATION" , 
         "HALLMARK_TNFA_SIGNALING_VIA_NFKB" , 
         'HALLMARK_INTERFERON_GAMMA_RESPONSE', 
         "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY", 
         "HALLMARK_NOTCH_SIGNALING", 
         "HALLMARK_APOPTOSIS", 
         "HALLMARK_HYPOXIA", 
         "BIOCARTA_GATA3_PATHWAY", 
         "GOBP_T_HELPER_2_CELL_DIFFERENTIATION",
         "GOBP_T_HELPER_17_CELL_DIFFERENTIATION", 
        "HALLMARK_CHOLESTEROL_HOMEOSTASIS", 
        "HALLMARK_ADIPOGENESIS", 
        "HALLMARK_MTORC1_SIGNALING", 
        "HALLMARK_PROTEIN_SECRETION", 
        "IL27_Upregulated_mouse"
        )

table(Ileum_filt$InfectionStatus, Ileum_filt$secondlevel)
unique(Ileum_filt$secondlevel)
meta <- Ileum_filt@meta.data
meta_subset <- meta %>%
 dplyr::select(InfectionStatus, secondlevel, FineCellType, SampleType,
         HALLMARK_INTERFERON_ALPHA_RESPONSE = HALLMARK_INTERFERON_ALPHA_RESPONSE, 
         HALLMARK_BILE_ACID_METABOLISM = HALLMARK_BILE_ACID_METABOLISM, 
         HALLMARK_GLYCOLYSIS =HALLMARK_GLYCOLYSIS, 
         HALLMARK_INFLAMMATORY_RESPONSE =HALLMARK_INFLAMMATORY_RESPONSE, 
         HALLMARK_WNT_BETA_CATENIN_SIGNALING = HALLMARK_WNT_BETA_CATENIN_SIGNALING, 
         HALLMARK_FATTY_ACID_METABOLISM = HALLMARK_FATTY_ACID_METABOLISM,
         HALLMARK_IL2_STAT5_SIGNALING = HALLMARK_IL2_STAT5_SIGNALING, 
         HALLMARK_TGF_BETA_SIGNALING = HALLMARK_TGF_BETA_SIGNALING,
         HALLMARK_IL6_JAK_STAT3_SIGNALING = HALLMARK_IL6_JAK_STAT3_SIGNALING, 
         HALLMARK_OXIDATIVE_PHOSPHORYLATION = HALLMARK_OXIDATIVE_PHOSPHORYLATION, 
         HALLMARK_TNFA_SIGNALING_VIA_NFKB = HALLMARK_TNFA_SIGNALING_VIA_NFKB, 
         HALLMARK_INTERFERON_GAMMA_RESPONSE = HALLMARK_INTERFERON_GAMMA_RESPONSE,
         HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY = HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY, 
         HALLMARK_NOTCH_SIGNALING = HALLMARK_NOTCH_SIGNALING,
         HALLMARK_APOPTOSIS = HALLMARK_APOPTOSIS, 
         HALLMARK_HYPOXIA = HALLMARK_HYPOXIA, 
         BIOCARTA_GATA3_PATHWAY = BIOCARTA_GATA3_PATHWAY, 
         GOBP_T_HELPER_2_CELL_DIFFERENTIATION = GOBP_T_HELPER_2_CELL_DIFFERENTIATION, 
         GOBP_T_HELPER_17_CELL_DIFFERENTIATION = GOBP_T_HELPER_17_CELL_DIFFERENTIATION, 
         HALLMARK_CHOLESTEROL_HOMEOSTASIS = HALLMARK_CHOLESTEROL_HOMEOSTASIS,
          HALLMARK_ADIPOGENESIS=HALLMARK_ADIPOGENESIS, 
        HALLMARK_MTORC1_SIGNALING =HALLMARK_MTORC1_SIGNALING, 
        HALLMARK_PROTEIN_SECRETION =HALLMARK_PROTEIN_SECRETION, 
        IL27_Upregulated_mouse = IL27_Upregulated_mouse
         )
# I ended up removing any cell types in which the naive condition did not have at least 10 cells - in order to preserve the interpretability of the data
meta_subset <- meta_subset[!meta_subset$secondlevel %in% c("Mesothelial Cells", "cDC1s", "Cycling ILCs", "Mast Cells", "Neutrophils"),]
meta_subset <- meta_subset[!meta_subset$SampleType %in% c("Duodenum Intestinal Epithelial Cells", "Duodenum Lamina Propria"),]
```

```{r}
library(rlang)
cellorder <- c("Enteric Glia",  "Monocytes" , "Macrophages" , "Plasmacytoid DCs" ,   "cDC2s" , 
               "ILC1s" , "ILC2s" ,  "ILC3s"  ,   "NK Cells"  , "NKT Cells" ,
               "Cycling T Cells",  "CD4 ab T Cells" , "Effector CD4 ab T Cells" , "CD8 ab T Cells" , "Effector CD8 ab T Cells" ,    "gd T Cells",
                "GC B Cells" ,  "B Cells", "Plasma Cells", 
                  "Stem Cells" , "Transit Amplifying Cells" ,  "Enterocytes" ,  "Goblet Cells"  ,  "Tuft Cells"   ,  "Paneth Cells"   ,  "Enteroendocrine Cells" ,
                "Lymphatic Endothelial Cells" , "Blood Endothelial Cells",   "Fibroblasts",         "Smooth Muscle Cells"   , "Stromal Cells"     )
infectionorder <- c("Yersinia", "SFB_YA", "Nippostrongylus", "MNV", "Cryptosporidium", "Candida", "Naive")

meta_subset <- meta_subset %>%
  mutate(across(all_of(pathways_interest), as.numeric))
heat <- list()
for (i in pathways_interest) {
  
  avg_scores <- meta_subset %>%
  group_by(InfectionStatus, secondlevel) %>%
  summarise(mean_score = mean(!!sym(i), na.rm = TRUE), .groups = "drop")
  
  
  # Scale mean_score *within* each cell type (secondlevel) across InfectionStatus
avg_scores_scaled <- avg_scores %>%
  group_by(secondlevel) %>%                      # scale within cell type
  mutate(scaled_score = as.numeric(scale(mean_score))) %>%
  ungroup()
avg_scores_scaled$secondlevel <- factor(avg_scores_scaled$secondlevel, levels = cellorder)
avg_scores_scaled$InfectionStatus <- factor(avg_scores_scaled$InfectionStatus, levels = infectionorder)

low <-min(avg_scores_scaled$mean_score)
max <-max(avg_scores_scaled$mean_score)
midpoint <-(low + max) /2

avg_scores$InfectionStatus <- factor(avg_scores_scaled$InfectionStatus, levels = infectionorder)
heat[[i]] <- ggplot(avg_scores_scaled, aes(x = secondlevel, y = InfectionStatus, fill = scaled_score)) +
  geom_tile(color =NA) +
  scale_fill_gradient2(
    low = "#1C75BC", mid = "grey", high = "#F7941D",
    midpoint = midpoint,
    name = "Scaled UCell Score"
  ) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  labs(title = paste0(i)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4, color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position = "right"
  )
}

heat[["HALLMARK_INTERFERON_ALPHA_RESPONSE"]]
heat[["HALLMARK_BILE_ACID_METABOLISM"]]
heat[["HALLMARK_GLYCOLYSIS"]]
heat[["HALLMARK_INFLAMMATORY_RESPONSE"]]
heat[["HALLMARK_WNT_BETA_CATENIN_SIGNALING"]]
heat[["HALLMARK_FATTY_ACID_METABOLISM"]]
heat[["HALLMARK_IL2_STAT5_SIGNALING"]]
heat[["HALLMARK_TGF_BETA_SIGNALING"]]
heat[["HALLMARK_IL6_JAK_STAT3_SIGNALING"]]
heat[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]
heat[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]]
heat[['HALLMARK_INTERFERON_GAMMA_RESPONSE']]
heat[["HALLMARK_APOPTOSIS"]]
heat[["HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY"]]
heat[['HALLMARK_NOTCH_SIGNALING']]
heat[['HALLMARK_HYPOXIA']]
heat[['BIOCARTA_GATA3_PATHWAY']]
heat[['GOBP_T_HELPER_2_CELL_DIFFERENTIATION']]
heat[['GOBP_T_HELPER_17_CELL_DIFFERENTIATION']]
heat[['HALLMARK_CHOLESTEROL_HOMEOSTASIS']]
heat[['HALLMARK_ADIPOGENESIS']]
heat[['HALLMARK_MTORC1_SIGNALING']]
heat[['HALLMARK_PROTEIN_SECRETION']]
heat[['IL27_Upregulated_mouse']]
      
```



```{r}
library(patchwork)
Wnt <- heat[["HALLMARK_WNT_BETA_CATENIN_SIGNALING"]]+ theme(axis.text.x = element_blank(), 
                                                            title = element_blank(), panel.grid = element_blank()
                                                            )

Gamma <- heat[['HALLMARK_INTERFERON_GAMMA_RESPONSE']]+theme( axis.text.x = element_blank(), title = element_blank())

Bile <-heat[["HALLMARK_BILE_ACID_METABOLISM"]]+ theme( title = element_blank())


(Gamma / Wnt / Bile) + plot_layout( heights = c(1, 1.1), nrow = 3) & 
  theme(legend.position = "right")
ggsave( "ssGSEA_Heatmap_Figure_v2.svg",  plot = last_plot() , device = NULL, 
        path = images, width = 10,  height = 7,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)
```

```{r}
#Fgiure 3C is derived from here
# Extract the information for Ifng to make a bubble plot showing that Crypto has increased expression
celltype_col <- "secondlevel"   
condition_col <- "InfectionStatus" 
sampletype <- "SampleType"
gene <- "Ifng"

expr <- FetchData(Ileum_filt, vars = c(gene, celltype_col, condition_col, sampletype))
expr <- expr[!expr$secondlevel %in% c("Mesothelial Cells", "cDC1s", "Cycling ILCs", "Mast Cells", "Neutrophils"),] # To match the Ucell scoring, remove cell types that have fewer than 10 cells in naive
expr <- expr[!expr$SampleType %in% c("Duodenum Intestinal Epithelial Cells", "Duodenum Lamina Propria"),] # Remove duodenum
unique(expr$SampleType) # check removal

bubble_data <- expr %>%
  group_by(!!sym(celltype_col), !!sym(condition_col)) %>%
  summarise(
    mean_expr = mean(.data[[gene]], na.rm = TRUE),                 
    pct_expr = sum(.data[[gene]] > 0) / n() * 100            
  ) %>%
  ungroup()

bubble_data$secondlevel <- factor(bubble_data$secondlevel, levels = rev(cellorder))
bubble_data$InfectionStatus <- factor(bubble_data$InfectionStatus, levels =rev(c("Yersinia", "SFB_YA", "Nippostrongylus", "MNV", "Cryptosporidium", "Candida", "Naive")))
ggplot(bubble_data, aes(x = secondlevel, y = InfectionStatus)) +
  geom_point(aes(size = pct_expr, fill = mean_expr), color = "black", shape = 21) +
  scale_size(range = c(1, 8)) +
  scale_fill_gradient(low = "lightgrey", high = "navyblue") +
  theme_minimal(base_size = 14) + theme(axis.title = element_blank(), panel.grid = element_blank(), 
                                        panel.grid.major = element_blank(), 
                                        panel.grid.minor = element_blank(),
                                        axis.text = element_text(color = "black"), 
                                        axis.text.x = element_text(angle = -90)) +
  labs(
    size = "% expressing",
    fill = "Mean expression"
  ) + coord_flip()
# Figure 3C
ggsave( "IfngExpression_allsamples_Fig2.svg",  plot = last_plot() , device = NULL, path = images, width = 10,  height = 11,  units = c("in"),  dpi = 600,  limitsize = TRUE,  bg = NULL)

```



```{r}
#Heatmaps of relevant pathways
#Save plots for figure 3 Figure 3D, Figure 3F, Figure 3H, Figure S4E

heat[["HALLMARK_INTERFERON_ALPHA_RESPONSE"]] + theme( title = element_blank())
ggsave("GSVA_hallmark_IFNa_AllSamples.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 11, height = 3.5, limitsize = TRUE, bg = NULL)

heat[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]] + theme( title = element_blank())
ggsave("GSVA_hallmark_IFNg_AllSamples.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 11, height = 3.5, limitsize = TRUE, bg = NULL)


heat[["GOBP_T_HELPER_17_CELL_DIFFERENTIATION"]] + theme( title = element_blank())
ggsave("BoBP_Th17_Ucell_Heatmap_AllSamples.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 11, height = 3.5, limitsize = TRUE, bg = NULL)

heat[["HALLMARK_TGF_BETA_SIGNALING"]] + theme( title = element_blank())
ggsave("Hallmark_TGFb_Ucell_Heatmap_AllSamples.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 11, height = 3.5, limitsize = TRUE, bg = NULL)

heat[["HALLMARK_HYPOXIA"]] + theme( title = element_blank())
ggsave("HHALLMARK_HYPOXIA_Ucell_Heatmap_AllSamples.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 11, height = 3.5, limitsize = TRUE, bg = NULL)
```



# ssGSEA Ucell ENrichments of the MLN ----
```{r}

hallmark <- data.table::fread("/path/to/analysis/directory/CSV/Hallmark_ssGSEA_UCell_MLN.csv")
rownames(hallmark) <- hallmark$cell_id
all(rownames(hallmark) %in% colnames(MLN_filt))
hallmark <- hallmark[,-1] # remove cell id column 
genesets_hallmark <- colnames(hallmark)
MLN_filt@meta.data <- MLN_filt@meta.data %>%
  dplyr::select(-starts_with("HALLMARK_"))
MLN_filt@assays[genesets_hallmark] <- NULL
MLN_filt <- AddMetaData(MLN_filt, metadata = hallmark)

Th2 <- data.table::fread("/path/to/analysis/directory/CSV/Th2_ssGSEA_Ucell_MLN.csv")
Th17 <- data.table::fread("/path/to/analysis/directory/CSV/Th17_ssGSEA_Ucell_MLN.csv")
rownames(Th2) <- Th2$cell_id
rownames(Th17) <- Th2$cell_id
all(rownames(Th2) %in% colnames(MLN_filt))
Th2 <- Th2[,-1] 
genesets_Th2 <- colnames(Th2)

MLN_filt <- AddMetaData(MLN_filt, metadata = Th2)

all(rownames(Th17) %in% colnames(MLN_filt))
Th17 <- Th17[, -1] 
Th2 <- data.table::fread("/path/to/analysis/directory/CSV/Th2_ssGSEA_Ucell_MLN.csv")
rownames(Th17) <- Th2$cell_id
genesets_Th17 <- colnames(Th17)

MLN_filt <- AddMetaData(MLN_filt, metadata = Th17)


```


```{r}
#Create a data frame of the score of selected Pathways 
pathways_interest  <- c("HALLMARK_INTERFERON_ALPHA_RESPONSE","HALLMARK_BILE_ACID_METABOLISM" , 
         "HALLMARK_GLYCOLYSIS" , 
         "HALLMARK_INFLAMMATORY_RESPONSE", 
         "HALLMARK_WNT_BETA_CATENIN_SIGNALING", 
         "HALLMARK_FATTY_ACID_METABOLISM" ,
         "HALLMARK_IL2_STAT5_SIGNALING" , 
         "HALLMARK_TGF_BETA_SIGNALING" ,
         "HALLMARK_IL6_JAK_STAT3_SIGNALING" , 
         "HALLMARK_OXIDATIVE_PHOSPHORYLATION" , 
         "HALLMARK_TNFA_SIGNALING_VIA_NFKB" , 
         'HALLMARK_INTERFERON_GAMMA_RESPONSE', 
         "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY", 
         "HALLMARK_NOTCH_SIGNALING", 
         "HALLMARK_APOPTOSIS", 
         "HALLMARK_HYPOXIA", 
        "HALLMARK_CHOLESTEROL_HOMEOSTASIS", 
        "HALLMARK_ADIPOGENESIS", 
        "HALLMARK_MTORC1_SIGNALING", 
        "HALLMARK_PROTEIN_SECRETION", 
         "BIOCARTA_GATA3_PATHWAY", 
         "GOBP_T_HELPER_2_CELL_DIFFERENTIATION",
         "GOBP_T_HELPER_17_CELL_DIFFERENTIATION" 
        )

table(MLN_filt$InfectionStatus, MLN_filt$secondlevel)
unique(MLN_filt$secondlevel)
meta <- MLN_filt@meta.data
meta_subset <- meta %>%
 dplyr::select(InfectionStatus, secondlevel, FineCellType,
         HALLMARK_INTERFERON_ALPHA_RESPONSE = HALLMARK_INTERFERON_ALPHA_RESPONSE, 
         HALLMARK_BILE_ACID_METABOLISM = HALLMARK_BILE_ACID_METABOLISM, 
         HALLMARK_GLYCOLYSIS =HALLMARK_GLYCOLYSIS, 
         HALLMARK_INFLAMMATORY_RESPONSE =HALLMARK_INFLAMMATORY_RESPONSE, 
         HALLMARK_WNT_BETA_CATENIN_SIGNALING = HALLMARK_WNT_BETA_CATENIN_SIGNALING, 
         HALLMARK_FATTY_ACID_METABOLISM = HALLMARK_FATTY_ACID_METABOLISM,
         HALLMARK_IL2_STAT5_SIGNALING = HALLMARK_IL2_STAT5_SIGNALING, 
         HALLMARK_TGF_BETA_SIGNALING = HALLMARK_TGF_BETA_SIGNALING,
         HALLMARK_IL6_JAK_STAT3_SIGNALING = HALLMARK_IL6_JAK_STAT3_SIGNALING, 
         HALLMARK_OXIDATIVE_PHOSPHORYLATION = HALLMARK_OXIDATIVE_PHOSPHORYLATION, 
         HALLMARK_TNFA_SIGNALING_VIA_NFKB = HALLMARK_TNFA_SIGNALING_VIA_NFKB, 
         HALLMARK_INTERFERON_GAMMA_RESPONSE = HALLMARK_INTERFERON_GAMMA_RESPONSE,
         HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY = HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY, 
         HALLMARK_NOTCH_SIGNALING = HALLMARK_NOTCH_SIGNALING,
         HALLMARK_APOPTOSIS = HALLMARK_APOPTOSIS, 
         HALLMARK_HYPOXIA = HALLMARK_HYPOXIA, 
         HALLMARK_CHOLESTEROL_HOMEOSTASIS = HALLMARK_CHOLESTEROL_HOMEOSTASIS,
          HALLMARK_ADIPOGENESIS=HALLMARK_ADIPOGENESIS, 
        HALLMARK_MTORC1_SIGNALING =HALLMARK_MTORC1_SIGNALING, 
        HALLMARK_PROTEIN_SECRETION =HALLMARK_PROTEIN_SECRETION, 
         BIOCARTA_GATA3_PATHWAY =BIOCARTA_GATA3_PATHWAY , 
         GOBP_T_HELPER_2_CELL_DIFFERENTIATION=GOBP_T_HELPER_2_CELL_DIFFERENTIATION,
         GOBP_T_HELPER_17_CELL_DIFFERENTIATION=GOBP_T_HELPER_17_CELL_DIFFERENTIATION 
         )
# I ended up removing any cell types in which the naive condition did not have at least 10 cells - in order to preserve the interpretability of the data
meta_subset <- meta_subset[!meta_subset$secondlevel %in% c("Mesothelial Cells", "Basophils", "Lymphatic Endothelial Cells", "Blood Endothelial Cells", "Fibroblasts", "Neutrophils"),]
```

```{r}
library(rlang)
unique(MLN_filt$secondlevel)
cellorder <- c( "Monocytes" , "Macrophages" ,"Plasmacytoid DCs"   ,   "cDC2s" ,  "cDC1s",
               "ILC1s" , "ILC2s" ,  "ILC3s"  ,   "NK Cells"  , "NKT Cells" ,
               "Cycling T Cells",  "CD4 ab T Cells" , "Effector CD4 ab T Cells" ,"Memory CD4 ab T Cells"  , "CD8 ab T Cells" , "Effector CD8 ab T Cells" ,    "gd T Cells",
                "GC B Cells" ,  "B Cells",    "Memory B cells"   ,"Plasma Cells"  )

infectionorder <- c("Yersinia", "SFB_YA", "Nippostrongylus", "MNV", "Cryptosporidium", "Candida", "Naive")

meta_subset <- meta_subset %>%
  mutate(across(all_of(pathways_interest), as.numeric))
heat_mln <- list()
for (i in pathways_interest) {
  
  avg_scores <- meta_subset %>%
  group_by(InfectionStatus, secondlevel) %>%
  summarise(mean_score = mean(!!sym(i), na.rm = TRUE), .groups = "drop")
  
  
  # Scale mean_score *within* each cell type (secondlevel) across InfectionStatus
avg_scores_scaled <- avg_scores %>%
  group_by(secondlevel) %>%                     
  mutate(scaled_score = as.numeric(scale(mean_score))) %>%
  ungroup()
avg_scores_scaled$secondlevel <- factor(avg_scores_scaled$secondlevel, levels = cellorder)
avg_scores_scaled$InfectionStatus <- factor(avg_scores_scaled$InfectionStatus, levels = infectionorder)

low <-min(avg_scores_scaled$mean_score)
max <-max(avg_scores_scaled$mean_score)
midpoint <-(low + max) /2

avg_scores$InfectionStatus <- factor(avg_scores_scaled$InfectionStatus, levels = infectionorder)
heat_mln[[i]] <- ggplot(avg_scores_scaled, aes(x = secondlevel, y = InfectionStatus, fill = scaled_score)) +
  geom_tile(color =NA) +
  scale_fill_gradient2(
     low = "#1C75BC", mid = "grey", high = "#F7941D",
    midpoint = midpoint,
    name = "Scaled UCell Score"
  ) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  labs(title = paste0(i)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4, color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position = "right"
  )
}

heat_mln[["HALLMARK_INTERFERON_ALPHA_RESPONSE"]]
heat_mln[["HALLMARK_BILE_ACID_METABOLISM"]]
heat_mln[["HALLMARK_GLYCOLYSIS"]]
heat_mln[["HALLMARK_INFLAMMATORY_RESPONSE"]]
heat_mln[["HALLMARK_WNT_BETA_CATENIN_SIGNALING"]]
heat_mln[["HALLMARK_FATTY_ACID_METABOLISM"]]
heat_mln[["HALLMARK_IL2_STAT5_SIGNALING"]]
heat_mln[["HALLMARK_TGF_BETA_SIGNALING"]]
heat_mln[["HALLMARK_IL6_JAK_STAT3_SIGNALING"]]
heat_mln[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]
heat_mln[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]]
heat_mln[['HALLMARK_INTERFERON_GAMMA_RESPONSE']]
heat_mln[["HALLMARK_APOPTOSIS"]]
heat_mln[["HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY"]]
heat_mln[['HALLMARK_NOTCH_SIGNALING']]
heat_mln[['HALLMARK_HYPOXIA']]
heat_mln[['BIOCARTA_GATA3_PATHWAY']]
heat_mln[['GOBP_T_HELPER_2_CELL_DIFFERENTIATION']]
heat_mln[['GOBP_T_HELPER_17_CELL_DIFFERENTIATION']]
heat_mln[['HALLMARK_CHOLESTEROL_HOMEOSTASIS']]
heat_mln[['HALLMARK_ADIPOGENESIS']]
heat_mln[['HALLMARK_MTORC1_SIGNALING']]
heat_mln[['HALLMARK_PROTEIN_SECRETION']]
heat_mln[['IL27_Upregulated_mouse']]
      
```

```{r}
##Figure 3E, Figure 3G 
heat_mln[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]] + theme( title = element_blank())
ggsave("HAllmark_IFNG_Ucell_Heatmap_AllSamples_MLN.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 8, height = 3.5, limitsize = TRUE, bg = NULL)


heat_mln[["GOBP_T_HELPER_17_CELL_DIFFERENTIATION"]] + theme( title = element_blank())
ggsave("GOBP_T_HELPER_17_CELL_Ucell_Heatmap_AllSamples_MLN.svg", plot = last_plot() ,device = NULL,
  path = images, scale = 1, width = 8, height = 3.5, limitsize = TRUE, bg = NULL)
```




#Supplement Graphing 
```{r}
#Figure S4A, Figure S4B, Figure S4C, Figure S4D
Correctcolors <- c(  "#B78CBA",   "#1170AA",  
  "#FCB13F",  
  "#60BFC1",  
  "#EF6F6A",  
    "#937860", 
  "#D17A00",   )
meta <- Ileum_filt_i@meta.data %>%
  dplyr::select(Mouse, InfectionStatus, secondlevel)

# Compute frequency per mouse and cell type
freq_data <- meta %>%
  group_by(InfectionStatus, secondlevel) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(InfectionStatus) %>%
  mutate(freq = n / sum(n)) %>%
  ungroup()
unique(freq_data$InfectionStatus)
freq_data <- freq_data %>%
  mutate(
    InfectionStatus = factor(
      InfectionStatus,
      levels = c("Naive", "Candida", "Cryptosporidium", "MNV",
                 "Nippostrongylus", "SFB_YA", "Yersinia")))
cell_types <- c("Monocytes", "Goblet Cells", "Neutrophils", "Tuft Cells")

plots <- lapply(cell_types, function(ct) {
  ggplot(filter(freq_data, secondlevel == ct),
         aes(x = InfectionStatus, y = freq, fill = InfectionStatus)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      title = ct,
      y = "Fraction of cells",
      x = "Infection"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_fill_manual(values = Correctcolors) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      
    )
})

# Combine all 4 plots in a grid
wrap_plots(plots, ncol = 2)

plots[[1]] + theme(axis.title.x = element_blank(), 
                   plot.title = element_blank(),
                   axis.text = element_text(color = "black"))
ggsave("Monocyte_frequency_Ileum_By_Infection.svg", plot = last_plot(), path = images, width = 6, height = 4)

plots[[2]] + theme(axis.title.x = element_blank(), 
                   plot.title = element_blank(),
                   axis.text = element_text(color = "black"))
ggsave("GobletCells_frequency_Ileum_By_Infection.svg", plot = last_plot(), path = images, width = 6, height = 4)

plots[[3]] + theme(axis.title.x = element_blank(), 
                   plot.title = element_blank(),
                   axis.text = element_text(color = "black"))
ggsave("Neutrophils_frequency_Ileum_By_Infection.svg", plot = last_plot(), path = images, width = 6, height = 4)

plots[[4]] + theme(axis.title.x = element_blank(), 
                   plot.title = element_blank(),
                   axis.text = element_text(color = "black"))
ggsave("Tuftcells_frequency_Ileum_By_Infection.svg", plot = last_plot(), path = images, width = 6, height = 4)

                   
                   
```
